---
title: "JC69 data with 12 species and 2000 sites" 
output:
  github_document:
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/shashanksule/Documents/info_theoretic_phylo/')
library("adegenet")
library("ape")
library("apTreeshape")
library("BoSSA")
library("diversitree")
library("pegas")
library("phangorn")
library("phylobase")
#library("phyloch")
library("seqinr")
library("readr")
source("utilities.R")
library("progress")
library("ggplot2")
library("phyclust")
library("TreeDist")
library("TreeTools")
library("dplyr")
library("parallel")
library("adephylo")
library("frequency")
library("plot3D")
```

We picked 105 samples from a set of 500 randomly generated JC69 sequences with the following parameters: 

```{r, eval = FALSE, echo=TRUE, include=FALSE}
setwd("/Users/shashanksule/Documents/info_theoretic_phylo/paml4.8")
n <- 500
trees <- list(n)
sequences <- list(n)
seeds <- list(n)
back_trees <- list(n)
for(i in 1:n){
  
  #Initialize the .dat file and store the seed
  seeds[[i]] <- who_dat(seqs = 12,
                sites = 2000, 
                model = 0,
                reps = 1,
                parameters = "1", 
                gamma = "1.0 0", 
                mutation = 5.5,
                equilibrium = "0.25 0.25 0.25 0.25",
                spit_seed = TRUE)
  
  #Simulate the trees and the sequences
  system("src/evolverRandomTree 5 MCbaseRTree.dat", show.output.on.console = FALSE) 
  #back_trees[[i]] <- ml_tree()
  trees[[i]] <- write.tree(read.nexus("mctrees.nex"))
  sequences[[i]] <- ReadCharacters("mc.nex")
}
```

These samples were found to have a scale ratio (minimum branch length to mutation) of more than 2.7%, the annealing threshold we picked for lowering sampling error. 

```{r, results='hide', echo =FALSE}
num_sequence <- 105
trees <- list(num_sequence)
agglomerative <- list(num_sequence)
agglomerative_asym <- list(num_sequence)
agglomerative_nj <- list(num_sequence)
divisive <- list(num_sequence)
divisive_asym <- list(num_sequence)
sequences <- list(num_sequence)
# agg_trees_nuc <- list(num_sequence)
# nj_trees_nuc <- list(num_sequence)

#read the original trees
for(i in c(1:num_sequence)){
  filename <- paste("./JC69_Jul28/Trees/tree", i, ".txt", sep = "")
  trees[[i]] <- readChar(filename, file.info(filename)$size)
}

#read the asymmetric divisive trees 
for(i in c(1:num_sequence)){
  filename <- paste("./JC69_Jul28/divisive_trees_sym/tree", i, ".txt", sep = "")
  divisive_asym[[i]] <- readChar(filename, file.info(filename)$size)
}
divisive_asym <- lapply(divisive_asym,  function(x) paste(x,";", sep = ""))

#read the symmetric divisive trees 
for(i in c(1:num_sequence)){
  filename <- paste("./JC69_Jul28/divisive_trees_sym/tree", i, ".txt", sep = "")
  divisive[[i]] <- readChar(filename, file.info(filename)$size)
}
divisive <- lapply(divisive,  function(x) paste(x,";", sep = ""))

#read the symmetric agglomerative trees
for(i in c(1:num_sequence)){
  filename <- paste("./JC69_Jul28/agglomerative_trees_sym/tree", i, ".txt", sep = "")
  agglomerative[[i]] <- readChar(filename, file.info(filename)$size)
}
agglomerative <- lapply(agglomerative,  function(x) paste(x,";", sep = ""))

#read the asymmetric agglomerative trees 
for(i in c(1:num_sequence)){
  filename <- paste("./JC69_Jul28/agglomerative_trees_sym/tree", i, ".txt", sep = "")
  agglomerative_asym[[i]] <- readChar(filename, file.info(filename)$size)
}
agglomerative_asym <- lapply(agglomerative_asym,  function(x) paste(x,";", sep = ""))

# #read the nj agglomerative trees
# for(i in c(1:num_sequence)){
#   filename <- paste("./JC69_Jul28/agglomerative_trees_nj/tree", i, ".txt", sep = "")
#   agglomerative_nj[[i]] <- readChar(filename, file.info(filename)$size)
# }
# agglomerative_nj <- lapply(agglomerative_nj,  function(x) paste(x,";", sep = ""))

#read the sequences 
for(i in c(1:num_sequence)){
  filename <- paste("./JC69_Jul28/Sequences/sequence", i, ".nex", sep = "")
  sequences[[i]] <- ReadCharacters(filename)
}

#make the neighbour joining trees 

nj_trees <- sequences %>%
            lapply(as.DNAbin) %>%
            lapply(dist.dna, model = "JC69") %>%
            lapply(nj) %>%
            lapply(write.tree)
```

1. Robinson Foulds distances 

```{r include = FALSE}
data <- data.frame(Original = as.character(trees), 
                   Agglomerative = as.character(agglomerative), 
                   Agglomerative_Asym = as.character(agglomerative_asym),
                   Agglomerative_NJ = as.character(agglomerative_nj),
                   Divisive = as.character(divisive), 
                   Divisive_Asym = as.character(divisive_asym),
                   NJ = as.character(nj_trees))
div_distance <- mapply(RobinsonFoulds,
       lapply(data$Divisive, function(x) read.tree(text = as.character(x))),
       lapply(data$Original, function(x) read.tree(text = as.character(x))))
agg_distance <- mapply(RobinsonFoulds,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$Agglomerative, function(x) read.tree(text = as.character(x))))
# agg_nj_distance <- mapply(RobinsonFoulds,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative_NJ, function(x) read.tree(text = as.character(x))))
nj_distance <- mapply(RobinsonFoulds,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$NJ, function(x) read.tree(text = as.character(x))))
```


```{r include = FALSE}
mean_div <- mean(div_distance)
mean_agg <- mean(agg_distance)
#mean_agg_nj <- mean(agg_nj_distance)
mean_nj <- mean(nj_distance)
prop_div <- mean(as.numeric(div_distance) == 0)
prop_agg <- mean(as.numeric(agg_distance) == 0)
#prop_agg_nj <- mean(as.numeric(agg_nj_distance) == 0)
prop_nj <- mean(as.numeric(nj_distance) == 0)

dist_data <- data.frame(Algorithm = c("Divisive","Agglomerative","NJ"),
                        Mean_RF = c(mean_div, mean_agg, mean_nj),
                        Props = c(prop_div, prop_agg, prop_nj))
colnames(dist_data) <- c("Algorithm", "Mean RF dist.", "Prop. of RF = 0")

```


```{r echo=FALSE, results='markup'}
show(dist_data)
```

```{r echo=FALSE, results='markup'}
layout(matrix(c(1,2,3,3), nrow = 2, ncol = 2))
hist(div_distance, main = "Hisogram of Divisive RF Distances")
hist(agg_distance, main = "Histogram of Agglomerative RF Distances")
hist(nj_distance, main = "Histogram of NJ RF Distances")
#hist(agg_nj_distance, main = "Histogram of NJ RF distances")
```

2. Branch length comparisons 

Among the trees computed by our algorithms, only the divisive trees had a few topologically identical ones 

```{r include=FALSE}
div_ids <- which(div_distance == 0)

og_tree <- read.tree(text = trees[[div_ids[1]]])
div_tree <- read.tree(text = divisive[[div_ids[1]]])
div_tree_rotat <- rotateConstr(div_tree, og_tree$tip.label)
og_tree_new <- root(og_tree, outgroup = "S1")
div_tree_new <- root(div_tree_rotat, outgroup = "S1")
div_dist_corr <- data.frame(Originals = og_tree_new$edge.length, Divisive = div_tree_new$edge.length)

for(i in div_ids[-1]){

og_tree <- read.tree(text = trees[[i]])
div_tree <- read.tree(text = divisive[[i]])
div_tree_rotat <- rotateConstr(div_tree, og_tree$tip.label)
og_tree_new <- root(og_tree, outgroup = "S1")
div_tree_new <- root(div_tree_rotat, outgroup = "S1")
div_dist_corr <- rbind(div_dist_corr, cbind(Originals = og_tree_new$edge.length, Divisive = div_tree_new$edge.length))
}
```

```{r, echo=FALSE}
ggplot(div_dist_corr, aes(x=Originals, y=Divisive))+
  geom_point(size = 4, alpha = 0.4) + 
  #xlim(c(0,6)) + 
  #ylim(c(0,1.0)) + 
  xlab("Branch Length of Original tree") + 
  ylab("Branch Length of NJ Tree (Bits per site)")
  #geom_smooth()
```

3. Tree properties (ultrametricity, additivity, rooting accuracy)

```{r include = FALSE}
div_root  <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$Divisive, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
div_asym_root <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$Divisive, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
agg_root <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$Agglomerative, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
# agg_asym_root <- mapply(all.equal,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative_Asym, function(x) read.tree(text = as.character(x))),
#        use.edge.length = FALSE) %>% as.numeric %>% mean
# agg_nj_root <- mapply(all.equal,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative_NJ, function(x) read.tree(text = as.character(x))),
#        use.edge.length = FALSE) %>% as.numeric %>% mean
nj_root <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$NJ, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
rooting <- c(1, div_root, div_asym_root, agg_root, nj_root)
```

```{r include = FALSE}
og_metricity <- data$Original %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean

div_metricity <- data$Divisive %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
div_asym_metricity <- data$Divisive_Asym %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
agg_metricity <- data$Agglomerative %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
# agg_asym_metricity <- data$Agglomerative_Asym %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
# agg_nj_metricity <- data$Agglomerative_NJ %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
nj_metricity <- data$NJ %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean

metricity <- c(1, div_metricity, div_asym_metricity, agg_metricity, nj_metricity)
```

```{r include = FALSE}
og_additivity <- data$Original %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
div_additivity <- data$Divisive %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
div_asym_additivity <- data$Divisive_Asym %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
agg_additivity <- data$Agglomerative %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
# agg_asym_additivity <- data$Agglomerative_Asym %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(is_additive) %>% as.numeric %>% mean
# agg_nj_additivity <- data$Agglomerative_NJ %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(is_additive) %>% as.numeric %>% mean
nj_additivity <- data$NJ %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean

additivity <- c(1, div_additivity, div_asym_additivity, agg_additivity, nj_additivity)
```

```{r include = FALSE}
Tree_data <- data.frame(Algorithm = c("Original", "Divisive", "Asymmetric Divisive", "Agglomerative", "NJ"),
                        Rooting = rooting,
                        Ultrametricity = metricity, 
                        Additivity = additivity)
```

```{r results = 'markup'}
print(Tree_data)
```

Aha! A few divisive trees actually get the rooting right. 