---
title: "TN93 data with 16 species and 1000 sites" 
output:
  github_document:
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/shashanksule/Documents/info_theoretic_phylo/')
library("adegenet")
library("ape")
library("apTreeshape")
library("BoSSA")
library("diversitree")
library("pegas")
library("phangorn")
library("phylobase")
#library("phyloch")
library("seqinr")
library("readr")
source("utilities.R")
library("progress")
library("ggplot2")
library("phyclust")
library("TreeDist")
library("TreeTools")
library("dplyr")
library("parallel")
library("adephylo")
library("frequency")
library("plot3D")
```

We generated a 100 sequences via the `evolver` with TN93 rate parameters based on the sequences simulated for Cummings et al 2003. 

```{r, echo=FALSE, results='hide', eval = TRUE, include=TRUE}
num_sequence <- 100
trees <- list(num_sequence)
agglomerative <- list(num_sequence)
agglomerative_asym <- list(num_sequence)
agglomerative_nj <- list(num_sequence)
divisive <- list(num_sequence)
divisive_asym <- list(num_sequence)
sequences <- list(num_sequence)
# agg_trees_nuc <- list(num_sequence)
# nj_trees_nuc <- list(num_sequence)

#read the original trees
for(i in c(1:num_sequence)){
  filename <- paste("./TN93_Jul21/Trees/tree", i, ".txt", sep = "")
  trees[[i]] <- readChar(filename, file.info(filename)$size)
}

#read the asymmetric divisive trees 
for(i in c(1:num_sequence)){
  filename <- paste("./TN93_Jul21/divisive_trees/tree", i, ".txt", sep = "")
  divisive_asym[[i]] <- readChar(filename, file.info(filename)$size)
}
divisive_asym <- lapply(divisive_asym,  function(x) paste(x,";", sep = ""))

#read the symmetric divisive trees 
for(i in c(1:num_sequence)){
  filename <- paste("./TN93_Jul21/divisive_trees/tree", i, ".txt", sep = "")
  divisive[[i]] <- readChar(filename, file.info(filename)$size)
}
divisive <- lapply(divisive,  function(x) paste(x,";", sep = ""))

# #read the symmetric agglomerative trees
# for(i in c(1:num_sequence)){
#   filename <- paste("./TN93_Jul21/agglomerative_trees/tree", i, ".txt", sep = "")
#   agglomerative[[i]] <- readChar(filename, file.info(filename)$size)
# }
# agglomerative <- lapply(agglomerative,  function(x) paste(x,";", sep = ""))
# 
# #read the asymmetric agglomerative trees 
# for(i in c(1:num_sequence)){
#   filename <- paste("./TN93_Jul21/agglomerative_trees/tree", i, ".txt", sep = "")
#   agglomerative_asym[[i]] <- readChar(filename, file.info(filename)$size)
# }
# agglomerative_asym <- lapply(agglomerative_asym,  function(x) paste(x,";", sep = ""))
# 
# #read the nj agglomerative trees
# for(i in c(1:num_sequence)){
#   filename <- paste("./TN93_Jul21/agglomerative_trees/tree", i, ".txt", sep = "")
#   agglomerative_nj[[i]] <- readChar(filename, file.info(filename)$size)
# }
# agglomerative_nj <- lapply(agglomerative_nj,  function(x) paste(x,";", sep = ""))

#read the sequences 
for(i in c(1:num_sequence)){
  filename <- paste("./TN93_Jul21/Sequences/sequence", i, ".nex", sep = "")
  sequences[[i]] <- ReadCharacters(filename)
}

#make the neighbour joining trees 

nj_trees <- sequences %>%
            lapply(as.DNAbin) %>%
            lapply(dist.dna, model = "TN93", base.freq = c(0.34, 0.28, 0.13, 0.25), 
                   gamma = 4) %>%
            lapply(nj) %>%
            lapply(write.tree)
```

1. Robinson-Foulds distances 

```{r, include=FALSE}
data <- data.frame(Original = as.character(trees), 
                   # Agglomerative = as.character(agglomerative), 
                   # Agglomerative_Asym = as.character(agglomerative_asym),
                   # Agglomerative_NJ = as.character(agglomerative_nj),
                   Divisive = as.character(divisive), 
                   Divisive_Asym = as.character(divisive_asym),
                   NJ = as.character(nj_trees))
div_distance <- mapply(RobinsonFoulds,
       lapply(data$Divisive, function(x) read.tree(text = as.character(x))),
       lapply(data$Original, function(x) read.tree(text = as.character(x))))
# agg_distance <- mapply(RobinsonFoulds,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative, function(x) read.tree(text = as.character(x))))
# agg_nj_distance <- mapply(RobinsonFoulds,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative_NJ, function(x) read.tree(text = as.character(x))))
nj_distance <- mapply(RobinsonFoulds,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$NJ, function(x) read.tree(text = as.character(x))))
```

```{r, include=FALSE}
mean_div <- mean(div_distance)
# mean_agg <- mean(agg_distance)
# mean_agg_nj <- mean(agg_nj_distance)
mean_nj <- mean(nj_distance)
prop_div <- mean(as.numeric(div_distance) == 0)
# prop_agg <- mean(as.numeric(agg_distance) == 0)
# prop_agg_nj <- mean(as.numeric(agg_nj_distance) == 0)
prop_nj <- mean(as.numeric(nj_distance) == 0)

dist_data <- data.frame(Algorithm = c("Divisive","NJ"),
                        Mean_RF = c(mean_div, mean_nj),
                        Props = c(prop_div, prop_nj))
colnames(dist_data) <- c("Algorithm", "Mean RF dist.", "Prop. of RF = 0")
```

```{r echo=FALSE, results='markup'}
show(dist_data)
```

```{r, include = FALSE}
layout(matrix(c(1,2), nrow = 1, ncol = 2))
hist(div_distance, main = "Hisogram of Divisive \nRF Distances from original", 
     w=2, xlab = "Robinson Foulds Distance")
hist(nj_distance, main = "Histogram of NJ \nRF Distances from original", w=2, xlab = "Robinson Foulds Distance")
```

2. Comparing branch lengths

There are, unfortunately, no divisive trees which had the same topology as the original ones. But we did have four such NJ trees. We plot the branch lengths of these trees against the branch lengths of the corresponding original trees: 

```{r, include=FALSE}
nj_ids <- which(nj_distance == 0)

og_tree <- read.tree(text = trees[[nj_ids[1]]])
nj_tree <- read.tree(text = nj_trees[[nj_ids[1]]])
nj_tree_rotat <- rotateConstr(nj_tree, og_tree$tip.label)
og_tree_new <- root(og_tree, outgroup = "S1")
nj_tree_new <- root(nj_tree_rotat, outgroup = "S1")
nj_dist_corr <- data.frame(Originals = og_tree_new$edge.length, NJ = nj_tree_new$edge.length)

for(i in nj_ids[-1]){

og_tree <- read.tree(text = trees[[i]])
nj_tree <- read.tree(text = nj_trees[[i]])
nj_tree_rotat <- rotateConstr(nj_tree, og_tree$tip.label)
og_tree_new <- root(og_tree, outgroup = "S1")
nj_tree_new <- root(nj_tree_rotat, outgroup = "S1")
nj_dist_corr <- rbind(nj_dist_corr, cbind(Originals = og_tree_new$edge.length, NJ = nj_tree_new$edge.length))
}
```

```{r echo = FALSE}
ggplot(nj_dist_corr, aes(x=Originals, y=NJ))+
  geom_point(size = 4, alpha = 0.4) + 
  #xlim(c(0,6)) + 
  #ylim(c(0,1.0)) + 
  xlab("Branch Length of Original tree") + 
  ylab("Branch Length of NJ Tree (Bits per site)")
  #geom_smooth()
```

3. Tree properties (ultrametricity, additivity, rooting accuracy)

```{r, include = FALSE}
div_root  <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$Divisive, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
div_asym_root <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$Divisive, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
# agg_root <- mapply(all.equal,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative, function(x) read.tree(text = as.character(x))),
#        use.edge.length = FALSE) %>% as.numeric %>% mean
# agg_asym_root <- mapply(all.equal,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative_Asym, function(x) read.tree(text = as.character(x))),
#        use.edge.length = FALSE) %>% as.numeric %>% mean
# agg_nj_root <- mapply(all.equal,
#        lapply(data$Original, function(x) read.tree(text = as.character(x))),
#        lapply(data$Agglomerative_NJ, function(x) read.tree(text = as.character(x))),
#        use.edge.length = FALSE) %>% as.numeric %>% mean
nj_root <- mapply(all.equal,
       lapply(data$Original, function(x) read.tree(text = as.character(x))),
       lapply(data$NJ, function(x) read.tree(text = as.character(x))),
       use.edge.length = FALSE) %>% as.numeric %>% mean
rooting <- c(1, div_root, div_asym_root, nj_root)
```

```{r include=FALSE}
og_metricity <- data$Original %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean

div_metricity <- data$Divisive %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
div_asym_metricity <- data$Divisive_Asym %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
# agg_metricity <- data$Agglomerative %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
# agg_asym_metricity <- data$Agglomerative_Asym %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
# agg_nj_metricity <- data$Agglomerative_NJ %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean
nj_metricity <- data$NJ %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(function(x) is.ultrametric(x, tol = 1e-06)) %>% as.numeric %>% mean

metricity <- c(1, div_metricity, div_asym_metricity, nj_metricity)
```

```{r include=FALSE}
og_additivity <- data$Original %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
div_additivity <- data$Divisive %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
div_asym_additivity <- data$Divisive_Asym %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean
# agg_additivity <- data$Agglomerative %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(is_additive) %>% as.numeric %>% mean
# agg_asym_additivity <- data$Agglomerative_Asym %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(is_additive) %>% as.numeric %>% mean
# agg_nj_additivity <- data$Agglomerative_NJ %>%
#                 lapply(function(x) read.tree(text = as.character(x))) %>%
#                 lapply(is_additive) %>% as.numeric %>% mean
nj_additivity <- data$NJ %>%
                lapply(function(x) read.tree(text = as.character(x))) %>%
                lapply(is_additive) %>% as.numeric %>% mean

additivity <- c(1, div_additivity, div_asym_additivity, nj_additivity)
```

```{r include=FALSE}
Tree_data <- data.frame(Algorithm = c("Original", "Divisive", "Asymmetric Divisive", "NJ"),
                        Rooting = rooting,
                        Ultrametricity = metricity, 
                        Additivity = additivity)
```

```{r results = 'markup'}
print(Tree_data)
```